---
title: "CH4_PoolDilution"
author: "Kendalynn A. Morris"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}

#load packages
library(PoolDilutionR)
library(dplyr)
library(ggplot2)
library(ggpmisc)
library(ggh4x)
library(lubridate)
library(readr)
library(scales)
library(tibble)
library(tidyr)
library(kableExtra)

theme_set(theme_minimal() + theme(text = element_text(size = 11)))
```

# Methane Pool Dilution

This code calculates gross methane production and consumption from pool dilution time series data. Methane concentrations and 13C content are measured from a closed system over time, the model optimizes production rate (P) and the first order rate constant (k), by minimizing error in the model-predicted total methane content, as well as the isotopic signature.The model optimizes rates by weighting information against the signal:noise ratio of concentration and 13C-signatures using measurement precision as well as the magnitude of change over time. The calculations used here are based on von Fischer and Hedin 2002, 10.1029/2001GB001448, with some modifications (see text below).  

Original code was written for data collected on a Picarro CRDS equipped with a SSIM2 unit.  

### Step 1: Read in data...  
```{r Read & Clean Data, echo = FALSE, message = TRUE}
# Get names of data files
files <- list.files("picarro/", pattern = "*.csv", full.names = TRUE)
# Helper function
read_file <- function(f) {
    message("Reading ", f)
    read_csv(f, col_types = "ccdcddcddddddddddddddddddddddccccc") %>%
        mutate(File = basename(f))
}
# Read in and pre-process data
lapply(files, read_file) %>%
    bind_rows() %>%
    # filter out standards and clean up columns
    filter(! id %in% c("SERC100ppm", "UMDzero", "SERCzero", "desert5000")) %>%
    mutate(Timestamp = mdy_hm(`Date/Time`, tz = "UTC")) %>%
    select(Timestamp, id, round, vol,
           `HR 12CH4 Mean`, `HR 12CH4 Std`,
           #HR stands for High Resolution
           `HR 13CH4 Mean`, `HR 13CH4 Std`,
           `HR Delta iCH4 Mean`, `HR Delta iCH4 Std`,
           `12CO2 Mean`, `12CO2 Std`,
           `13CO2 Mean`, `13CO2 Std`,
           `Delta 13CO2 Mean`, `Delta 13CO2 Std`,
           notes) %>%
    # calculate elapsed time for each sample
    arrange(id, round) %>%
    group_by(id) %>%
    mutate(time_days = difftime(Timestamp, min(Timestamp),
                                units = "days"),
           time_days = as.numeric(time_days)) ->
    incdat_raw

summary(incdat_raw)
```

```{r Set Constants, echo=FALSE}
VOL_ML <- 130  # total volume of jar
```

### Step 2: Convert units and calculate time steps and normalization factors...  
```{r Convert Units, echo=TRUE}
incdat_raw %>%
    # Volume of jar = 130 ml or 0.130 L, 1 ppm = 0.001 ml/L
    # The Picarro measures 20 ml, but 10 ml injected (see vol);
    # So ppm is a 1:1 dilution (1 part sample to 1 part dilutant)
    # therefore ppm to ml = ppm * 0.001 * VOL_ML/1000 * 2
    mutate(cal12CH4ml = `HR 12CH4 Mean` * 0.001 * VOL_ML/1000 * 2 * 1000,
           cal13CH4ml = `HR 13CH4 Mean` * 0.001 * VOL_ML/1000 * 2 * 1000,
           # for each 10 ml sample from 130 ml jar,
           # 10 ml of zero air injected
           # 12 parts sample to 1 part dilutant (13 parts total)
           # remaining gas in jar is diluted 12:1
           cal12CH4ml = if_else(round != "T0", cal12CH4ml * 1.083, cal12CH4ml),
           cal13CH4ml = if_else(round != "T0", cal13CH4ml * 1.083, cal13CH4ml),
           # calculate atom percent (AP) of 13C methane in sample over time
           AP_obs = cal13CH4ml / (cal12CH4ml + cal13CH4ml) * 100) %>% 
  group_by(id) -> incdat

#now we generate sample specific scaling factors for feeding into a cost function
#see vFH2002 Eqs. 12, 13, & 14
#need standard deviation of isotopes and mass across all measurements for each sample -> sd_obs
#and standard deviation of the analytical precision aka
#instrument stdev for those observations -> sd_inst
# incdat %>% 
#     group_by(id) %>% 
#     mutate(m_prec = 0.001,
#            ap_prec = 1
    #this is now done within the cost_function
    #        sd_obsDelta = sd(`HR Delta iCH4 Mean`),
    #        sd_instDelta = sd(sdD),
    #        #sdD is the instrument stdev for the delta measurement
    #        sd_obsMass = sd(`HR 12CH4 Mean`),
    #        sd_instMass = sd(sdC),
    #        #sdC is the instrument stdev for the 12C mass measurement
   #        ) -> incdat
```

#### Input Data

```{r Input Data Visualization, echo=FALSE}
facet_labels <- c(
    AP_obs = "Atom% Observed",
    cal12CH4ml = "12C Methane (mL)",
    cal13CH4ml = "13C Methane (mL)")

incdat %>%
    filter(id %in% c("2","31","52","61","64","71")) %>%  
    mutate(id_numeric = as.factor(id)) %>%
    pivot_longer(cols = c(cal12CH4ml, cal13CH4ml, AP_obs)) %>%
    ggplot(aes(round, value, group = id, color = id)) +
    scale_color_discrete("Sample ID") +
    geom_point() + geom_line() + ylab("") + xlab("\n Timestep") +
    ggtitle("Example Input Data") +
    facet_wrap( ~ name, scales = "free",
                labeller = labeller(name = facet_labels)) -> p
print(p)
```

### Step 3: Atom Precent Prediction  

Prediction function\
time: vector of time values, numeric (e.g. days); first should be zero\
m0: amount of total methane at time zero\
n0: amount of labeled methane at time zero\
P: production rate of total methane, unit gas/unit time\
k: first-order rate constant for methane consumption, 1/unit time\
Returns a data frame with mt, nt, and AP (atom percent) predictions for each time step\
All combined, this is Eq. 11 from vFH2002 with a few modifications...\
1. Added x 100 so that the left part is correctly a percent (per their Appendix A)\
2. Removed addition of AP_P from Eq. 10 & 11, see notes in ReadMe\
3. Expanded Eq. 9 (and numerator in Eq. 11) to include both production and consumption terms for 13C methane (ie., naturally occuring heavy methane)\

### Step 4: Cost Function  

Called by optim()\
params: named vector holding optimizer-assigned values for P and k\
time: vector of time values, numeric (e.g. days); first should be zero\
m: observed total methane values, same length as time\
n: observed labeled methane values, same length as time\
Nd: normalization factor for delta values, see Eq. 12\
Nm: normalization factor for methane mass, see Eq. 13\
Returns the sum of squares between predicted and observed m and AP\

### Step 5: optimize P and k, then calculate consumption rate (C)  

```{r Do the thing!, echo=TRUE, message = FALSE, warning = FALSE, results='hide'}





###Here lies the problem.





pk_results <- list()
incdat_out <- list()
#all_predictions <- list()

for(i in unique(incdat$id)) {
    message("------------------- ", i)
    # Isolate this sample's data
    incdat %>%
        filter(id == i) %>%
        select(id, round, vol, time_days,
               cal12CH4ml, cal13CH4ml,
               AP_obs) ->
        dat
    
    # Let optim() try different values for P and k until it finds best fit to data
    #params <- list("P0" = 10) # k0 is auto-estimated
    result <- pdr_optimize(time = dat$time_days,
                          m = dat$cal12CH4ml + dat$cal13CH4ml,
                          n = dat$cal13CH4ml,
                          P = 0.1,
                          pool = "CH4",
                          m_prec = 0.001,
                          ap_prec = 1)#,
                          #params_to_optimize = c("P", "k"))
    
    # Save progress details separately so they don't print below
    progress_detail <- result$progress
    result$progress <- NULL

    message("Optimizer solution:")
    print(result)
    P <- result$par["P"]
    pk_results[[i]] <- tibble(P = P,
                              k = result$par["k"],
                              k0 = result$initial_par["k0"],
                              convergence = result$convergence,
                              message = result$message)
    
    # Predict based on the optimized parameters
    pred <- ap_prediction(time = dat$time_days,
                          m0 = dat$cal12CH4ml[1] + dat$cal13CH4ml[1],
                          n0 = dat$cal13CH4ml[1],
                          P = P,
                          k = result$par["k"],
                          pool = "CH4")
    dat <- bind_cols(dat, pred)
    
    # # Predict based on ALL the models that were tried
    # x <- split(progress_detail, seq_len(nrow(progress_detail)))
    # all_preds <- lapply(x, FUN = function(x) {
    #     y1<- data.frame(P = x$P[1],
    #                     k = x$k[1],
    #                     time = seq(min(dat$time_days), max(dat$time_days), length.out = 20))
    #     y2 <- ap_prediction(time = y1$time,
    #                     m0 = dat$cal12CH4ml[1] + dat$cal13CH4ml[1],
    #                     n0 = dat$cal13CH4ml[1],
    #                     P = x$P[1],
    #                     k = x$k[1],
    #                     pool = "CH4")
    #     cbind(y1, y2)
    # })
    # all_predictions[[i]] <- bind_rows(all_preds)

    # Calculate implied consumption (ml) based on predictions
    # Equation 4: dm/dt = P - C, so C = P - dm/dt
    total_methane <- dat$cal12CH4ml + dat$cal13CH4ml
    change_methane <- c(0, diff(total_methane))
    change_time <- c(0, diff(dat$time_days))
    dat$Pt <- P * change_time #P is ml/day
    #amount of methane produced at time (t) of this incubation, a volume in mL
    dat$Ct <- dat$Pt - change_methane
    #amount of methane consumed at time (t) of this incubation, a volume in mL

    incdat_out[[i]] <- dat
}
```

```{r Wrap it up!, echo=FALSE, message = FALSE}
pk_results <- bind_rows(pk_results, .id = "id")
incdat_out <- bind_rows(incdat_out)

incdat_out %>%
    # compute correlation between predictions and observations
    group_by(id) %>%
    summarise(m_cor = cor(cal12CH4ml + cal13CH4ml, mt),
              ap_cor = cor(AP_obs, AP_pred)) ->
    performance_summary

performance_summary %>%
    right_join(pk_results, by = "id") ->
    pk_results
print(pk_results)

pk_results %>%
    right_join(incdat_out, by = "id") ->
    incdat_out
```

# Results  

```{r Summarize Output}
summary(incdat_out)
```

# Summary Plots  

```{r Summary Plots, echo = FALSE}
# ----- Plot AP results -----
EXAMPLES <- c("2","31","52","61","64","71")
incdat_out %>%
    filter(id %in% EXAMPLES) %>%
    ggplot(aes(time_days, AP_obs, color = as.factor(id))) +
    geom_point(aes(shape = ""), size = 3.75) +
    #geom_line(data = filter(all_predictions, id %in% EXAMPLES),
              #aes(time, AP_pred, group = paste(id, P, k)), color = "grey", linetype = 2) +
    geom_line(aes(time_days, AP_pred, group = id, linetype = ""), linewidth = 1.25) +
    scale_linetype_manual(name = "Prediction",
                          values = "dotted") +
    scale_shape_manual(name = "Observations",
                       values = 20) +
    scale_color_discrete(guide = "none") +
    facet_wrap(~as.numeric(id), scales = "free") +
    xlab("\n Timestep \n") + ylab("\n (13C-CH4/Total CH4) x 100 \n") +
    ggtitle("\n Atom% 13C \n") + theme(legend.position = "bottom")


# ----- Plot total methane results -----
incdat_out %>%
    filter(id %in% EXAMPLES) %>%
    ggplot(aes(time_days, cal12CH4ml + cal13CH4ml, color = as.factor(id))) +
    geom_point(aes(shape = ""), size = 3.75) +
    # geom_line(data = filter(all_predictions, id %in% EXAMPLES),
    #           aes(time, mt, group = paste(id, P, k)), color = "grey", linetype = 2) +
    geom_line(aes(y = mt, group = id, linetype = ""), linewidth = 1.25) +
    scale_linetype_manual(name = "Prediction",
                          values = "dotted") +
    scale_shape_manual(name = "Observations",
                       values = 20) +
    scale_color_discrete(guide = "none") +
    facet_wrap(~as.numeric(id), scales = "free") +
    xlab("\n Timestep \n") + ylab("\n Volume (mL) \n") +
    ggtitle("\n Total Methane \n") + theme(legend.position = "bottom")
```


```{r Fit Trends, echo = FALSE}
# ----- Visualize data, coloring by fit -----
incdat_out %>%
    filter(id %in% EXAMPLES) %>%  
    pivot_longer(cols = c(cal12CH4ml, cal13CH4ml, AP_obs)) %>%
    ggplot(aes(round, value, group = id, color = ap_cor)) +
    geom_point() + geom_line() + xlab("\n Timestep") +
    ylab("") + scale_color_continuous(name = "Goodness of Fit") +
    facet_wrap( ~ name, scales = "free",
                labeller = labeller(name = facet_labels)) +
    ggtitle("Fit Trends") + theme(legend.position = "bottom")

```


```{r Results Table, echo = FALSE}
incdat_out %>%
    filter(id %in% c("2","31","52","61","64","71"),
           round == "T1") %>%
    select(id, Pt, k, Ct, AP_pred, AP_obs) %>%
    mutate(across(where(is.numeric), ~ num(., digits = 2))) -> results

kable(results,
      caption = "<center><strong>Gross Rates</strong></center>",
      escape = FALSE,
      col.names = NULL,
      align = "c",
      format = "html") %>%
    kable_styling(font_size = 12) %>% 
  add_header_above(c("Sample ID", "P mL/day", "k 1/day", "C mL/day", "Atom% Predicted", "Atom% Observed"))
```
