---
title: "GrossFlux"
author: "Kendalynn A. Morris"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
#load packages
library(PoolDilutionR)
library(dplyr)
library(generics)
library(ggplot2)
library(ggpmisc)
library(ggh4x)
library(lubridate)
library(readr)
library(scales)
library(tibble)
library(tidyr)
library(kableExtra)

theme_set(theme_minimal() + theme(text = element_text(size = 11)))

# 'Pretty n' function to round a numeric value and print that # of digits
pn <- function(x, n) {
  formatC(round(unlist(x), n),
          digits = n, format = "f")
}

# 'Clean p value' function to pretty-print p value(s), specifically
pclean <- function(x, digits = 3, printP = TRUE) {
  x <- as.vector(x)
  ltstring <- paste0("< 0.", paste(rep("0", digits - 1), 
                                   collapse = ""), "1")
  valstring <- ifelse(x < 10 ^ -digits, 
                      ltstring, pn(x, digits))
  if(printP) {
    paste("P", ifelse(x < 10 ^ -digits, 
                      valstring, paste("=", valstring)))
  } else {
    valstring
  }
}
```

# Methane Pool Dilution

Here are the results and analysis for the gross methane flux portion of Morris et al 2023b. Fluxes are calculated using PoolDilutionR, data was collected using a Picarro CRDS equipped with a SSIM2 unit.  

### Step 1: Read in data...  

```{r Read & Clean Data, echo = FALSE, message = TRUE}
# Get names of data files
files <- list.files("picarro/", pattern = "*.csv", full.names = TRUE)
# Helper function
read_file <- function(f) {
    message("Reading ", f)
    read_csv(f, col_types = "ccdcddcddddddddddddddddddddddccccc") %>%
        mutate(File = basename(f))
}
# Read in and pre-process data
lapply(files, read_file) %>%
    bind_rows() %>%
    # filter out standards and clean up columns
    filter(! id %in% c("SERC100ppm", "UMDzero", "SERCzero", "desert5000")) %>%
    mutate(Timestamp = mdy_hm(`Date/Time`, tz = "UTC")) %>%
    select(Timestamp, id, round, vol,
           `HR 12CH4 Mean`, `HR 12CH4 Std`,
           #HR stands for High Resolution
           `HR 13CH4 Mean`, `HR 13CH4 Std`,
           `HR Delta iCH4 Mean`, `HR Delta iCH4 Std`,
           `12CO2 Mean`, `12CO2 Std`,
           `13CO2 Mean`, `13CO2 Std`,
           `Delta 13CO2 Mean`, `Delta 13CO2 Std`,
           notes) %>%
    # calculate elapsed time for each sample
    arrange(id, round) %>%
    group_by(id) %>%
    mutate(time_days = difftime(Timestamp, min(Timestamp),
                                units = "days"),
           time_days = as.numeric(time_days)) ->
    incdat_raw

summary(incdat_raw[,1:10])
```

### Step 2: Convert units and calculate time steps and normalization factors...

```{r Convert Units, echo=TRUE}
VOL_ML <- 130  # total volume of jar

incdat_raw %>%
    # Volume of jar = 130 ml or 0.130 L, 1 ppm = 0.001 ml/L
    # The Picarro measures 20 ml, but 10 ml injected (see vol);
    # So ppm is a 1:1 dilution (1 part sample to 1 part dilutant)
    # therefore ppm to ml = ppm * 0.001 * VOL_ML/1000 * 2
    mutate(cal12CH4ml = `HR 12CH4 Mean` * 0.001 * VOL_ML/1000 * 2 * 1000,
           cal13CH4ml = `HR 13CH4 Mean` * 0.001 * VOL_ML/1000 * 2 * 1000,
           # for each 10 ml sample from 130 ml jar,
           # 10 ml of zero air injected
           # 12 parts sample to 1 part dilutant (13 parts total)
           # remaining gas in jar is diluted 12:1
           cal12CH4ml = if_else(round != "T0", cal12CH4ml * 1.083, cal12CH4ml),
           cal13CH4ml = if_else(round != "T0", cal13CH4ml * 1.083, cal13CH4ml),
           # calculate atom percent (AP) of 13C methane in sample over time
           AP_obs = cal13CH4ml / (cal12CH4ml + cal13CH4ml) * 100) %>% 
  group_by(id) -> incdat
```

```{r Do the thing!, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE, results='hide'}
pk_results <- list()
incdat_out <- list()
all_predictions <- list()

for(i in unique(incdat$id)) {
    message("------------------- ", i)
    # Isolate this sample's data
    incdat %>%
        filter(id == i) %>%
        select(id, round, vol, time_days,
               cal12CH4ml, cal13CH4ml,
               AP_obs) ->
        dat
    
    # Let optim() try different values for P and k until it finds best fit to data
    # Here we provide starting values for P (0.1), k (k0), and P_frac and k_frac are set to the methane default
    result <- pdr_optimize(time = dat$time_days,
                          m = dat$cal12CH4ml + dat$cal13CH4ml,
                          n = dat$cal13CH4ml,
                          P = 0.1,
                          pool = "CH4",
                          m_prec = 0.5,
                          ap_prec = 0.01,
                          include_progress = TRUE)
    
    # Save progress details separately so they don't print below
    progress_detail <- result$progress
    result$progress <- NULL

    #message("Optimizer solution:")
    #print(result)
    P <- result$par["P"]
    pk_results[[i]] <- tibble(P = P,
                              k = result$par["k"],
                              k0 = result$initial_par["k"],
                              convergence = result$convergence,
                              message = result$message)
    
    # Predict based on the optimized parameters
    pred <- pdr_predict(time = dat$time_days,
                          m0 = dat$cal12CH4ml[1] + dat$cal13CH4ml[1],
                          n0 = dat$cal13CH4ml[1],
                          P = P,
                          k = result$par["k"],
                          pool = "CH4")
    dat <- bind_cols(dat, pred)
    
    # Predict based on ALL the models that were tried
    
    x <- split(progress_detail, seq_len(nrow(progress_detail)))
    all_preds <- lapply(x, FUN = function(x) {
      y1<- data.frame(P = x$P[1],
                      k = x$k[1],
                      time = seq(min(dat$time_days), max(dat$time_days), length.out = 20))
      y2 <- pdr_predict(time = y1$time,
                          m0 = dat$cal12CH4ml[1] + dat$cal13CH4ml[1],
                          n0 = dat$cal13CH4ml[1],
                          P = x$P[1],
                          k = x$k[1],
                          pool = "CH4")
      cbind(y1, y2)
    })
    all_predictions[[i]] <- bind_rows(all_preds)

    # Calculate implied consumption (ml) based on predictions
    # Equation 4: dm/dt = P - C, so C = P - dm/dt
    total_methane <- dat$cal12CH4ml + dat$cal13CH4ml
    change_methane <- c(0, diff(total_methane))
    change_time <- c(0, diff(dat$time_days))
    dat$Pt <- P * change_time #P is ml/day
    #amount of methane produced at time (t) of this incubation, a volume in mL
    dat$Ct <- dat$Pt - change_methane
    #amount of methane consumed at time (t) of this incubation, a volume in mL

    incdat_out[[i]] <- dat
}
```

```{r Wrap it up!, echo=FALSE, message = FALSE}
pk_results <- bind_rows(pk_results, .id = "id")
incdat_out <- bind_rows(incdat_out)
all_predictions <- bind_rows(all_predictions, .id = "id")

incdat_out %>%
    # compute correlation between predictions and observations
    group_by(id) %>%
    summarise(m_cor = cor(cal12CH4ml + cal13CH4ml, mt),
              ap_cor = cor(AP_obs, AP_pred)) ->
    performance_summary

performance_summary %>%
    right_join(pk_results, by = "id") ->
    pk_results
print(pk_results)

pk_results %>%
    right_join(incdat_out, by = "id") ->
    incdat_out
```

# Results  

```{r Summarize Output}
summary(incdat_out)
```

# Summary Plots  

```{r Summary Plots, echo = FALSE, fig.height = 10, fig.width = 12}
# ----- Plot AP results -----
incdat_out %>%
  ggplot(aes(time_days, AP_obs, color = as.factor(id))) +
  geom_point(aes(shape = ""), size = 3.75) +
  geom_line(data = all_predictions,
            aes(time, AP_pred, group = paste(id, P, k)), color = "grey", linetype = 2) +
  geom_line(aes(y = AP_pred, group = id, linetype = ""),
            linewidth = 1.25) +
  scale_linetype_manual(name = "Prediction",
                        values = "dotted") +
  scale_shape_manual(name = "Observations",
                     values = 20) +
  scale_color_discrete(guide = "none") +
  facet_wrap(~as.numeric(id), scales = "free") +
  xlab("\n Timestep \n") + ylab("\n (13C-CH4/Total CH4) x 100 \n") +
  ggtitle("\n Atom% 13C \n") + theme(legend.position = "bottom")


# ----- Plot total methane results -----
incdat_out %>%
  ggplot(aes(time_days, cal12CH4ml + cal13CH4ml, color = as.factor(id))) +
  geom_point(aes(shape = ""), size = 3.75) +
  geom_line(data = all_predictions,
            aes(time, mt, group = paste(id, P, k)), color = "grey", linetype = 2) +
  geom_line(aes(y = mt, group = id, linetype = ""),
            linewidth = 1.25) +
  scale_linetype_manual(name = "Prediction",
                        values = "dotted") +
  scale_shape_manual(name = "Observations",
                     values = 20) +
  scale_color_discrete(guide = "none") +
  facet_wrap(~as.numeric(id), scales = "free") +
  xlab("\n Timestep \n") + ylab("\n Volume (mL) \n") +
  ggtitle("\n Total Methane \n") + theme(legend.position = "bottom")
```

# Normalize by Soil Mass (g)

```{r Incubation Soil Data}
#read in soil moisture data
smDat <- read_csv("field-measurements/July22_soilmoisture.csv",
                  col_types = "ccdddcc")
#combine lab data (no flux data yet)
labDat <- merge(smDat, incdat_out, by = "id")
#calculate soil dry mass and fresh water mass
labDat$mass <- labDat$jdry - labDat$jempty
labDat$sm <- (labDat$jfresh - labDat$jdry)/labDat$mass

#ml to umols, n = (PV/RT)*10^6
#R = 82.0574 mL * atm/K*mol
#T = 297.15 K, estimated lab temp
#P = 1 atm
#V = Pt or Ct

labDat$umolP <- (labDat$Pt/(82.0574*297.15)) * 10^6
labDat$umolC <- (labDat$Ct/(82.0574*297.15)) * 10^6

#convert to rate per g dry soil per day
labDat$P_rate <- (labDat$umolP/labDat$time_days)/labDat$mass
labDat$C_rate <- (labDat$umolC/labDat$time_days)/labDat$mass

labDat %>%
  filter(round == "T4", Location == "g_low") %>%
  select(Origin, sm, P_rate, C_rate) %>%
  pivot_longer(sm:C_rate, names_to = "var") %>%
  group_by(Origin, var) %>%
  summarize(across(everything(), list(Mean = mean, SD = sd),
                   .names = "{.fn}"),
            .groups = "drop") %>%
  pivot_wider(names_from = "var",
              values_from = c(Mean, SD),
              names_glue = "{var}_{.value}") -> Lowland
print(Lowland)

labDat %>%
  filter(round == "T1", Location == "g_up") %>%
  select(Origin, sm, P_rate, C_rate) %>%
  pivot_longer(sm:C_rate, names_to = "var") %>%
  group_by(Origin, var) %>%
  summarize(across(everything(), list(Mean = mean, SD = sd),
                   .names = "{.fn}"),
            .groups = "drop") %>%
  pivot_wider(names_from = "var",
              values_from = c(Mean, SD),
              names_glue = "{var}_{.value}") -> Upland
print(Upland)

```

# Initial Analysis

```{r normalization of gross fluxes, echo = FALSE, results='hide'}

#select T4 as the representative flux for all analyses. Justification: at the end of the incubation head space gas composition will most closely resemble the atmosphere.
labDat %>%
  filter(round == "T4") -> grossFlux

hist(grossFlux$sm)
hist(grossFlux$P_rate)
hist(grossFlux$C_rate)

require(rcompanion)
transformTukey(grossFlux$P_rate,
               plotit = TRUE, verbose = TRUE)
transformTukey(grossFlux$C_rate,
               plotit = TRUE, verbose = TRUE)

hist(grossFlux$P_rate^0.025)
hist(-1 * grossFlux$C_rate^-0.075)
```

Soil samples come from 5 different origins (lowland, upland, midslope, upstream, and midstream) and are situated in two physical locations (lowland, upland). Does soil origin, current location, or an interaction between the two explain variation in gross methane production or consumption?

```{r production}
require(car)
P_transplant <- aov(P_rate^0.025 ~ Location*Origin,
                  data = grossFlux)
Anova(P_transplant)
Ptrans <- unlist(Anova(P_transplant))
Ptrans_loc <- Ptrans["Pr(>F)1"]
P_moisture <- aov(P_rate^0.025 ~ sm,
                  data = grossFlux)
Anova(P_moisture)
Pmois <- unlist(Anova(P_moisture))
Pmois_sm <- Pmois["Pr(>F)1"]
```
There's some evidence that location influences methane production (`r pclean(Ptrans_loc, 3)`), and equally good evidence that soil moisture does (`r pclean(Pmois_sm, 3)`.

```{r consumption}
require(car)
C_transplant <- aov(-1 * C_rate^-0.075 ~ Location*Origin,
                  data = grossFlux)
Anova(C_transplant)
C_moisture <- aov(C_rate ~ sm,
                  data = grossFlux)
Anova(C_moisture)
```

There's strong evidence that location influences methane consumption, and even stronger evidence that soil moisture does.

```{r moisture}
require(car)
sm_transplant <- aov(sm ~ Location*Origin,
                  data = grossFlux)
Anova(sm_transplant)
```

Location strongly influences soil moisture values, which have a clear influence on methane gross rates. In order to better test the influence of soil origin, we separate the two locations and test the effect of origin without the interaction of current location.

# Alternative Models

```{r Within Upland, echo = FALSE, results='hide'}
grossFlux %>%
  filter(Location == "g_up") -> UpFlux

hist(UpFlux$sm)
hist(UpFlux$P_rate)
hist(UpFlux$C_rate)

transformTukey(UpFlux$P_rate,
               plotit = TRUE, verbose = TRUE)
transformTukey(UpFlux$C_rate,
               plotit = TRUE, verbose = TRUE)

hist(UpFlux$P_rate^0.025)
hist(-1 * UpFlux$C_rate^-0.725)

upP_transplant <- aov(P_rate^0.025 ~ Origin,
                  data = UpFlux)
Anova(upP_transplant)
upC_transplant <- aov(-1 * C_rate^-0.725 ~ Origin,
                  data = UpFlux)
Anova(upC_transplant)

upP_moisture <- aov(P_rate^0.025 ~ sm,
                  data = UpFlux)
Anova(upP_moisture)
upC_moisture <- aov(-1 * C_rate^-0.725 ~ sm,
                  data = UpFlux)
Anova(upC_moisture)
```

This makes a compelling argument that soil origin has no influences on gross methane production or consumption in the upland, rather the variation that is present in gross consumption is best explained by soil moisture. 

```{r, Within Lowland, echo = FALSE, results='hide'}
grossFlux %>%
  filter(Location == "g_low") -> LowFlux

hist(LowFlux$sm)
hist(LowFlux$P_rate)
hist(LowFlux$C_rate)

transformTukey(LowFlux$P_rate,
               plotit = TRUE, verbose = TRUE)
transformTukey(LowFlux$C_rate,
               plotit = TRUE, verbose = TRUE)

hist(-1 * LowFlux$P_rate^-0.325)
hist(-1 * LowFlux$C_rate^-0.025)

LowP_transplant <- aov(-1* P_rate^0.325 ~ Origin,
                  data = LowFlux)
Anova(LowP_transplant)
LowC_transplant <- aov(-1 * C_rate^-0.025 ~ Origin,
                  data = LowFlux)
Anova(LowC_transplant)

LowP_moisture <- aov(-1* P_rate^0.325 ~ sm,
                  data = LowFlux)
Anova(LowP_moisture)
LowC_moisture <- aov(-1 * C_rate^-0.025 ~ sm,
                  data = LowFlux)
Anova(LowC_moisture)
```

And same for the lowland location.

# Summary Plot of Gross Rates

```{r, Summary Graph of Analytes, echo = FALSE results='hide'}
#summary of production
Olabs <- c("lowland", "midslope", "upland", "midstream", "upstream")
Llabs <- c("lowland", "upland")
labellies <- c('g_low' = "lowland",
               'g_mid' = "midslope",
               'g_up' = "upland",
               'midstream' = "midstream",
               'upstream' = "upstream")
grossFlux %>%
  select(Location, Origin, P_rate, C_rate, sm) %>%
  pivot_longer(P_rate:sm, names_to = "vars") %>%
ggplot(., aes(Location, value, fill = Origin)) +
    scale_x_discrete(labels = c("lowland",
                                "upland")) +
    scale_fill_discrete(labels = Olabs) +
    geom_boxplot() + facet_wrap(.~vars, scales = "free")

```

# Distribution of Gross Rates

```{r net of gross}
grossFlux$diff <- grossFlux$P_rate - grossFlux$C_rate

ggplot(grossFlux, aes(diff, fill = Location)) +
  scale_fill_discrete(labels = Llabs) +
  geom_histogram(binwidth = 0.5, color = "black")

```

# Linear Regression with Soil Moisture


```{r Consumption vs SM}
ggplot(grossFlux, aes(sm, C_rate)) +
  geom_point(aes(color = Location, shape = Location), size = 3) +
  scale_color_manual(values = c("#FF33CC", "#00CC66"),
                     labels = Llabs) +
  scale_shape_discrete(labels = Llabs) +
    geom_smooth(method = lm, formula = y ~ x, se = FALSE, size = 0.5) +
  labs(x=  "Soil Moisture
       (g\uFE63\u00b9 dry soil)",
       y= "Consumption FIX UNIT") +
    stat_poly_eq(formula = y ~ x,
                 aes(label = paste(..eq.label..,
                                   ..rr.label..,
                                   sep = "~~~"))) +
  theme_bw(base_size = 15) +
  theme(plot.margin = margin(t = 20, r = 20, b = 20, l = 20))
```

```{r Production vs SM}
ggplot(grossFlux, aes(sm, P_rate)) +
  geom_point(aes(color = Location, shape = Location), size = 3) +
  scale_color_manual(values = c("#FF33CC", "#00CC66"),
                     labels = Llabs) +
  scale_shape_discrete(labels = Llabs) +
    geom_smooth(method = lm, formula = y ~ x, se = FALSE, size = 0.5) +
  labs(x=  "Soil Moisture
       (g\uFE63\u00b9 dry soil)",
       y= "Production FIX UNIT (d\uFE63\u00b9)") +
    stat_poly_eq(formula = y ~ x,
                 aes(label = paste(..eq.label..,
                                   ..rr.label..,
                                   sep = "~~~"))) +
  theme_bw(base_size = 15) +
  theme(plot.margin = margin(t = 20, r = 20, b = 20, l = 20))
```
